```{r echo=FALSE, eval=TRUE, results='hide',include=FALSE}
library(TMB)
library(mgcv)
library(Matrix)
source("../utils.R")
source("simpleExample.R")
```

## **Simple example ** 
In this example you learn:

* How to set up and fit a linear regression model in TMB.
* Basis functionality of TMB

***
#### Description of the practical situation
In this example we model nitrogen ratios (NL) in a theets of a whale as a linear function of its age. We assume $\text{NL}_{age} \sim N(\mu_{age},\sigma^2)$, were: $\mu_{age} = \beta_0 + \beta_{age}age$.
This example is found on page 27 in Zuur (2006).
</details>

<details> <summary> Summary of data.</summary>
The data is the same used on page 27 in Zuur (2006).
```{r,echo=FALSE, comment=""}
TN = read.table("TeethNitrogen.txt",header = T)
head(TN,5)
```
</details>


***

#### Full code for example
<details> <summary> R code</summary>
```{r,echo=FALSE, comment=""}
include_source("simpleExample.R", linesToInclud=c(1:21))
```
</details>
<details> <summary> C++ code</summary>
```{r,echo=FALSE, comment=""}
include_source("simpleExample.cpp", linesToInclud=c(1:999))
```
</details>



***
#### C++ code step-by-step
<details>
<summary>Read data from R</summary>
```c++
DATA_VECTOR(NL);  // Response
DATA_VECTOR(x);   // Covariate

```
</details>

<details>
<summary>Read parameters from R</summary>
```c++
PARAMETER(beta0);
PARAMETER(beta1);
PARAMETER(logSigma);
```
</details>

<details>
<summary>Transform some of the parameters. </summary>
```c++
Type sigma= exp(logSigma);
```
</details>

<details>
<summary>Define the objective function (where we will later store) 
$\text{nll}=-\log P(\text{NL}|\pmb{\beta}, \sigma)$. </summary>
```c++
Type nll=0; //negative log-likelihood
```
</details>

<details>
<summary>Calculates the regression line </summary>
```c++
vector<Type> mu = beta0 + x*beta1;
```
* This is a vector calculation. `vector<Type>` is the vector type.
</details>

<details>
<summary>Calculates the likelihood contribution from the observations </summary>
```c++
nll -= sum(dnorm(NL,mu,sigma,true));
```
* `dnorm(NL,mu,sigma,true)` is normal density (vectorized) in R-style notation.
* Argument `true` returns log-density (rather than density).
* `sum()` is adding the vector as an alternative to using a loop.
</details>

<details>
<summary>Return the objective function </summary>
```c++
return nll;
```
</details>

***

#### R code step-by-step
<details>  
<summary>Include the TMB library</summary>
```r
library(TMB)
```
</details>

<details>  
<summary>Compile and load the C++ code</summary>
```r
compile("simpleExample.cpp")  # Compile the C++ file
dyn.load("simpleExample")     # Link C++ file into R session
```
</details>

<details>  
<summary>Read data</summary>
```r
TN = read.table("TeethNitrogen.txt",header = T)
```
</details>

<details>  
<summary> Construct the data list which is given to TMB (C++)</summary>
```r
data = list()
data$NL = TN$X15N[TN$Tooth=="Moby"]
data$x = TN$Age[TN$Tooth=="Moby"]
```
</details>

<details>  
<summary> Construct the parameter list with initial values which is given to TMB (C++)</summary>
```r
parameters = list(
beta0 = 0,
beta1 = 0,
logSigma = 0
)
```
</details>

<details>  
<summary> Fit the model</summary>
```r
obj = MakeADFun(data,parameters,DLL = "simpleExample")  # Link data and model
opt = nlminb(obj$par,obj$fn, obj$gr)   # Estimate parameters
rep = sdreport(obj)                    # Evaluate standard deviations
```
</details>


***
#### Exercise


1. 
<details>  
<summary> Go trough the code and confirm that you understand what is happening. Plot the fitted regression line together with the data.
</summary>
    ```{r, fig.height=4, fig.width=4}
        rep$par.fixed
        plot(data$x,data$NL,xla = "Age", ylab ="NL")
        abline(a = rep$par.fixed[1], b = rep$par.fixed[2])
    ```
</details>

2. 
<details>  
<summary> Evaluate the objective function value 
at $\beta_0=0.3$, $\beta_{age}=0.4$ and $\log(\sigma)=0.2$.
</summary>
```r
obj$fn(c(0.3,0.4,0.2))
```
</details>

3. 
<details>  
<summary> Evaluate the gradient (derivative with respect to
the parameters) of the objective function at $\beta_0=0.3$, $\beta_{age}=0.4$ and $\log(\sigma)=0.2$.
</summary>
```r
obj$gr(c(0.3,0.4,0.2))
```
</details>
