```{r echo=FALSE, eval=TRUE, results='hide',include=FALSE}
library(TMB)
source("../utils.R")
```


## **Turning parameters on-and-off in the estimation** 
In this example you learn to:

* Turn off parameters in the estimation (fix at initial value)
* Contrain parameters to have the same value
* Put upper and lower bounds  on parameters in the function minimizer

***
#### Description of the practical situation
We use a simple two-factor regression
model (simulated data) with the first factor 
having 4 levels (fixed effect)
and the second factor having 10 levels (random effect):
$$
  y_k = \beta_{i_k} + u_{j_k} + e_k, \quad k=1,\dots,n
$$
where $u_j \sim N(0,\tau^2)$, $j=1,\ldots,10$ and $e_k  \sim  N(0,\sigma^2)$, $k=1,\ldots,n$.

The purpose of the case study is to show how the `map`
argument of `MakeADFun()` can be used to control which parameters
should be estimated (or fixed at their initial values). This
is useful in model comparison, where we want to compare
the full model against various reduced models.

Familiarity with random effects modelling in TMB is assumed in this case study.

***

#### Full code for example
<details> <summary> R code</summary>
```{r,echo=FALSE, comment=""}
include_source("map_example.R", linesToInclud=c(1:999))
```
</details>
<details> <summary> C++ code</summary>
```{r,echo=FALSE, comment=""}
include_source("map_example.cpp", linesToInclud=c(1:999))
```
</details>


***
#### Exercises

<details> <summary>Read about the `map` argument under
the Details section of the `?MakeADFun` help page.</summary>

Key points:

* Parameter entries with NAs in the factor are fixed.
* Parameter entries with equal factor level are collected to a common value.

```r
map=list(beta=as.factor(c(NA,NA,1,1)))
```

</details>

<details>
<summary>Impose identical fixed effects: $\beta_1=\beta_2=\beta_3=\beta_4$. Verify that `opt$par` only has a single `beta` component. </summary>
```{r,echo=FALSE, comment=""}
include_source("map_example.R", linesToInclud=c(33:35))
```
</details>

<details>
  <summary>Two groups of fixed effects: $\beta_1=\beta_2$ and $\beta_3=\beta_4$.</summary>
```{r,echo=FALSE, comment=""}
include_source("map_example.R", linesToInclud=c(38:40))
```

* $\beta_1$ and $\beta_2$ are estimated together as a single value, and similarly for $\beta_3$ and $\beta_4$.

* `opt$par` now should contain two `beta` components, one for
each group.

</details>

<details>
  <summary>Fix a parameter at its initial value: $\beta_1=0$.</summary>
```{r,echo=FALSE, comment=""}
include_source("map_example.R", linesToInclud=c(43:45))
```

* `opt$par` now should contain three `beta` components, corresponding
to $\beta_2$, $\beta_3$ $\beta_4$.

</details>

<details>
  <summary>Fix all $\beta$'s at their its initial values: 
  $\beta_1=\beta_2=\beta_3=\beta_4=0$.</summary>
```{r,echo=FALSE, comment=""}
include_source("map_example.R", linesToInclud=c(48:50))
```

* `opt$par`  shouldn't contain any `beta` component.

</details>

<details>
  <summary>Remove the random effects part of the model. </summary>

* Need to remove both $u$ vector and associated standard 
  deviation $\tau$.
```{r,echo=FALSE, comment=""}
include_source("map_example.R", linesToInclud=c(53:55))
```

* Comment: Could have removed `random="u"` from parameter list.

</details>

<details>
  <summary>Put bounds on the parameters (of the full model):
  $-10\leq \beta_i \leq 10$, $0\leq \sigma \leq 10$ , $0\leq \tau \leq 10$.  </summary>

```{r,echo=FALSE, comment=""}
include_source("map_example.R", linesToInclud=c(58:64))
```

</details>

<details>
  <summary>Remove the random effects (from full model), and modify the bounds correspondingly.</summary>
```{r,echo=FALSE, comment=""}
include_source("map_example.R", linesToInclud=c(67:70))
```

</details>


***
