```{r echo=FALSE, eval=TRUE, results='hide',include=FALSE}
library(TMB)
library(mgcv)
library(Matrix)
source("../utils.R")
source("pSplines.R")
```


## **GAM model based on P-splines**
In this example you learn:

* How to estimate a model with P-splines (penalized B-splines) in TMB.
* How to set up the the spline part of the model in the `mgcv` package. 
* How to extract standard errors of a spline in TMB.
* How to input "complex datastructures" from R to C++ via _sparse matrices_.

***
#### Description of the practical situation
Let $Y(d)$ be the biodiversity, as given by the variable `Richness` in the data. We assume that $Y(d) \sim N(\mu,\sigma^2)$, were 
$$\mu = \beta_0 +  f_1(rock) +f_2(litter) + f_3(soil) + f_4(fallPrec) + f_5(sprTmax).$$
Here, $f_1,\ldots,f_5$ are B splines, modelled as function of covariates `rock`, `litter`,... (see data summary below).
See page 63 in Zuur (2009) for a similar example with use of the same data.
<details> <summary> "Penalizing the splines": how to get an appropriate level of smoothness?</summary>
We let the coefficients of the $i$th spline have a multivariate normal distribution with precision matrix $\lambda_i \pmb{S}_i$, where $\lambda_i$ is a penalization parameter and $\pmb{S}_i$ is a penalization matrix extracted from the R-package `mgcv`. The matrices $\pmb{S}_i$ are singular, and hence the multivariate normal distributions are degenerate. However, when bringing in data through the likelihood, the (posterior) precission matrix becomes non-singular. See [Wood (2017)](https://books.google.no/books?id=JTkkDwAAQBAJ&printsec=frontcover&dq=generalized+additive+models+wood+2017&hl=no&sa=X&ved=0ahUKEwjp4o2sicHbAhWEFZoKHaHYBQ0Q6AEIJzAA#v=onepage&q=generalized%20additive%20models%20wood%202017&f=false) for details.
</details>

<details> <summary> Summary of the data.</summary>
```{r,echo=FALSE, comment=""}
head(Vegetation,5)
```
</details>

***

#### Full code for example
<details> <summary> R code</summary>
```{r,echo=FALSE, comment=""}
include_source("pSplines.R", linesToInclud=c(1:3, 10:999))
```
</details>
<details> <summary> C++ code</summary>
```{r,echo=FALSE, comment=""}
include_source("pSplines.cpp", linesToInclud=c(1:999))
```
</details>

***

#### C++ code step-by-step
<details>  
  <summary>Read data from R</summary>
```c++
    DATA_VECTOR(Y);
    DATA_MATRIX(X);         // Covariates as columns
    DATA_SPARSE_MATRIX(S);  // diag(S1,S2,S3,S4,S5), but not storing 
                            // off-diagonal zeros
    DATA_VECTOR(Sdims);     // Dimensions of S1,S2,S3,S4,S5
    DATA_SPARSE_MATRIX(designMatrixForReport);
```
</details>
<details>
  <summary>Read parameters from R</summary>
```c++
    PARAMETER(beta0);
    PARAMETER_VECTOR(beta);
    PARAMETER_VECTOR(log_lambda);
    PARAMETER(log_sigma);
```
</details>
<details>

<summary>Transform the standard deviation, $\sigma$, and the penalize parameter, $\lambda$, from log scale</summary>
```c++
    Type sigma = exp(log_sigma);
    vector<Type> lambda = exp(log_lambda);
```
</details>
<details>

<summary>Define the objective function</summary>
```c++
    Type nll = 0.0;
```
</details>
<details>

  <summary>Calculate the the likelihood contribution from the spline (the penalilzation)</summary>
  The negative log likelihood contribution from each spline is $0.5 \text{log} \lambda_i m_i -  \lambda_i \pmb{\beta}_i \pmb{S}_i \pmb{\beta}_i^T$, were $m_i$ is the number of regression parameters for the $i$th spline. 
```c++
    vector<Type> S_beta = S*beta;
    nll -= 0.5*(log_lambda*Sdims).sum();  
    int counter = 0;
    for(int i=0;i<Sdims.size(); i++){
      for(int j=0;j<Sdims(i); j++){
        nll -= -0.5*lambda(i)*beta(counter)*S_beta(counter);
        counter++;
      }
    }
```
</details>
<details>

  <summary>Calculate the liear predictor, $\mu$.   </summary>
```c++
    vector<Type> mu(Y.size());
    mu = beta0 + X*beta;
```
</details>
<details>

<summary>Calculate the negative log likelihood $\sum_{i=1}^{n}\log P(Y|\sigma,\text{splines})$</summary>
```c++
    for(int i=0; i<Y.size(); i++){
      nll -= dnorm(Y(i), mu(i), sigma, true);
    }
```
</details>
<details>

<summary>Return the objective function </summary>
```c++
  return nll;
```
</details>


***
#### Exercise

1. There are 5 penalty matrices (S-matrices) in this example that are sent
to the .cpp code jointly as a blockdiagonal sparse matrix `S_combined`.
Understand this mechanism by looking at the following miniature example:
<details>  
  <summary>R code to be run line by line</summary>
```r
  M1 = matrix(1:4,2,2)               # ordinary 2x2 matrix
  M2 = matrix(5:9,2,2)               # ordinary 2x2 matrix
  M_block_diag = .bdiag(list(M1,M2)) # Block diagonal (sparse) matrix
  data.class(M_block_diag)           # Check data.class
  print(M_block_diag)                # dots means 0 value
```
</details>


2. Plot the estimated splines with 95% confidence intervalls.
<details>
  <summary>Solution</summary>
    First we define a design matrix for the splines we want to plot:
    ```r
      #R code
      ROCK=min(Vegetation$ROCK):max(Vegetation$ROCK)
      LITTER=min(Vegetation$LITTER):max(Vegetation$LITTER)
      BARESOIL=min(Vegetation$BARESOIL):max(Vegetation$BARESOIL)
      FallPrec=min(Vegetation$FallPrec):max(Vegetation$FallPrec)
      SprTmax=min(Vegetation$SprTmax):max(Vegetation$SprTmax)
      
      rockReport = PredictMat(gam_setup$smooth[[1]],data = data.frame(ROCK))
      litterReport = PredictMat(gam_setup$smooth[[2]],data = data.frame(LITTER)) 
      soilReport = PredictMat(gam_setup$smooth[[3]],data = data.frame(BARESOIL)) 
      fallReport = PredictMat(gam_setup$smooth[[4]],data = data.frame(FallPrec)) 
      sprReport = PredictMat(gam_setup$smooth[[5]],data = data.frame(SprTmax))
      
      designMatrixForReport = list(rockReport,litterReport,soilReport,fallReport,sprReport)
      data$designMatrixForReport = .bdiag(designMatrixForReport) 
    ```
    
    Then we estimate those splines in TMB when we estimate the model:
    ```c++
        //c++ code
        DATA_SPARSE_MATRIX(designMatrixForReport); //place were the data is included
        
        vector<Type> splineForReport = designMatrixForReport*beta; //place just before return nll
        ADREPORT(splineForReport);
  
    ```
    Then we extract the estimated splines with standard errors, and plot the splines:
  
    ```{r, fig.height=7, fig.width=7}
      #R code
      muSpline = rep$value[names(rep$value)=="splineForReport"]
      sdSpline<-rep$sd[names(rep$value)=="splineForReport"]
      
      par(mfrow=c(2,3))
      start = 1
      stop = start + length(ROCK) -1
      plot(ROCK, muSpline[start:stop], lty=1,type = 'l',ylim = c(-6,5),ylab = "f(rock)",main = "Spline for ROCK")
      lines(ROCK, muSpline[start:stop]- 1.96*sdSpline[start:stop], lty=2)
      lines(ROCK, muSpline[start:stop]+ 1.96*sdSpline[start:stop], lty=2)
      abline(h = 0)
      
      start = stop +1
      stop = start+ length(LITTER)-1
      plot(LITTER, muSpline[start:stop], lty=1,type = 'l',ylim = c(-6,5),ylab = "f(litter)",main = "Spline for LITTER")
      lines(LITTER, muSpline[start:stop]- 1.96*sdSpline[start:stop], lty=2)
      lines(LITTER, muSpline[start:stop]+ 1.96*sdSpline[start:stop], lty=2)
      abline(h = 0)
      
      start = stop +1
      stop = start+ length(BARESOIL)-1
      plot(BARESOIL, muSpline[start:stop], lty=1,type = 'l',ylim = c(-6,5),ylab = "f(soil)",main = "Spline for BARESOIL")
      lines(BARESOIL, muSpline[start:stop]- 1.96*sdSpline[start:stop], lty=2)
      lines(BARESOIL, muSpline[start:stop]+ 1.96*sdSpline[start:stop], lty=2)
      abline(h = 0)
      
      start = stop +1
      stop = start+ length(FallPrec)-1
      plot(FallPrec, muSpline[start:stop], lty=1,type = 'l',ylim = c(-6,5),ylab = "f(fallPrec)",main = "Spline for FallPrec")
      lines(FallPrec, muSpline[start:stop]- 1.96*sdSpline[start:stop], lty=2)
      lines(FallPrec, muSpline[start:stop]+ 1.96*sdSpline[start:stop], lty=2)
      abline(h = 0)
      
      start = stop +1
      stop = start+ length(SprTmax)-1
      plot(SprTmax, muSpline[start:stop], lty=1,type = 'l',ylim = c(-6,5),ylab = "f(sprTMax)",main = "Spline for SprTmax")
      lines(SprTmax, muSpline[start:stop]- 1.96*sdSpline[start:stop], lty=2)
      lines(SprTmax, muSpline[start:stop]+ 1.96*sdSpline[start:stop], lty=2)
      abline(h = 0)
    ```
</details>


</details>

