---
title: "Observations"
author: "Hans Skaug, Jens Wahl, Olav Breivik"
date: "12 juni 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(TMB)
source("../../utils.R")
```

## **Observations**

So far we have looked at the three main processes that drives the SAM model, namely the recruitment $N_{1,y}$, survival $N_{a,y}, a > 1$ and the fishing mortality $F_{a,y}$. We have treated them as processes that we observe, but in SAM the are all unobserved processes.
We will in this exercise focus on the observational equation for 

* Catch-at-age $C_{a,y}$
* Survey index-at-age $I_{a,y}$

Even thought unobserved in the full SAM model, we will in this exercise assume that we have observed the recruitment, survival and fishing mortality. 

***

### Catch-at-age

The catch-at-age is predicted by the catch equation: 

\[
  \hat{C}_{a,y} = \frac{F_{a,y}}{F_{a,y} + M_{a,y}} N_{a,y} (1 - e^{- F_{a,y} - M_{a,y}}),
\]

where $N$ is the stocksize, $F$ is the fishing mortality and $M$ is natural mortality. This equation is based on the assumption that the suvival cohort size will develop over time accoring to 
\[
  \frac{dN}{dt} = -(F + M)N,
\]
which has the solution $N = N_0 e^{-F-M}$, meaning that the dead population size is $N_0(1 - e^{-F-M})$. This means that the equation $\frac{F_{a,y}}{F_{a,y} + M_{a,y}} N_{a,y} (1 - e^{- F_{a,y} - M_{a,y}})$ tells us the fraction of the dead population caused by fishing.   

For convinience, we will define the total mortality as $Z_{a,y} = F_{a,y} + M_{a,y}$. Taking the logarithm and assuming gaussian noise, we can write over model as 

\[
  \log C_{a,y} = \log F_{a,y} - \log Z_{a,y} + \log N_{a,y} + log(1 - e^{-Z_{a,y}}) + \epsilon_{a,y},
\]
where $\epsilon_{a,y} \sim N(0, \sigma^2)$. 

### The data 

* The data list in ```Cobs.RData``` contains a vector ```Cobs``` of the catch observations. It is a
vector not a matrix.

* In addition the data list contains a matrix ```aux``` with three columns ```(year, fleet, age)```.
The $i$th row in this matrix contains the information for the $i$th element in the Cobs vector.

* Further the data list contains information on ```F, N, M, minYear```, and ```minAge```. 

*** 


### Full code for example 
<details> 
  <summary> R code </summary> 
```{r,echo=FALSE, comment=""}
include_source("Cobs.R", linesToInclud=c(1:999))
```
</details>

<details> <summary> C++ code</summary>
```{r,echo=FALSE, comment=""}
include_source("Cobs.cpp", linesToInclud=c(1:999))
```
</details>


*** 

## Including survey fleets

We will now extend the model we just implemented to include survey fleets, but also handle missing data. This will be done by adding them as random effects and estimating them. 

The survey indices are modelles as: 
\[
  \log I_{y}^{(s)} = \log ( Q_a^{(s)} e^{-(F_{a,y} + M_{a,y})day^{(s)}/365} N_{a,y}) + \epsilon^s_{a,y},  
\]

where 

* $Q_a^{(s)}$ is the proportionality coefficient for age $a$ and survey $s$. This will be estimated. 
* $day^{(s)}/365$ is the time into the year the survey $s$ is conducted. 

***

### The data

* The data list in `allfleets.RData` contains a vector `obs` with all observations (catches
and surveys).
* In addition the data list contains a matrix ```aux``` with three columns ```(year, fleet, age)```.
The $i$th row in this matrix contains the information for the $i$th element in the Cobs vector.
* Data also contain a vector ```fleetTypes``` with three elements. ‘0’ indicates a catch fleet
and ‘2’ indicates a survey fleet.
* Finally the data contains a vector `sampleTimes` with three elements, which is the $day^{(s)}/365$
values (only used for the survey fleets).

### Handling missing observations

Some of the observations in the `obs` vector are missing `NA`. We could just remove these from the likelihood, but a smarter way to deal with them is to treat them as random effects. On the R-side we add the missing values to the parameter list: 
```{r, eval = F}
missing = numeric(sum(is.na(allfleets$obs)))
```

We can make a helper function for dealing with missing values. This can be pasted right below `#include<TMB.hpp>`: 

```c++
// Function that checks if data is missing 
// Returns logical true or false
template<class Type> 
bool isNA(Type x){
  return R_IsNA(asDouble(x));
}
```

and impute the missing values: 

```c++
// count missing values 
int index_is_missing = 0; 
for(int i = 0; i < n; i++){
 // Estimate missing data
 if(isNA(obs(i))){
   obs(i) = exp(missing(index_is_missing++));
 }
}
```

## Configuring parameters 

In this data set we have 3 fleets and 9 ages, but not all fleets have all ages. We only need to focus on fleet 2 and 3, as they are the only survey fleets. If we define an integer data matrix like 

```{r, eval = F}
allfleets$keyQ <- rbind(rep(NA, 9), 
                        c(NA, 0, 1, 2, 3, 4, 5, 6, NA),
                        c(7, 8, 9, 10, 11, 12, NA, NA, NA))
```

and a parameter like 

```{r, eval = F}
logQ = numeric(max(allfleets$keyQ, na.rm = T) + 1)
```

we can us the table to look up which model parameter we should use for a given fleet $f$ and a age $a$. This can be done like: 

```c++ 
case 2:
  logPred(i) = logQ(keyQ(fleet, age)) + log(N(year, age)) - sampleTimes(fleet) * Z; 
break; 
```

### Exercise 

1. 
<details> 
  <summary> Extend the privous model for catch-at-age to include surveys. </summary> 
