---
title: "Observations"
author: "Hans Skaug, Jens Wahl, Olav Breivik"
date: "12 juni 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(TMB)
source("../../utils.R")
```

## **Observations**

So far we have looked at the three main processes that drive the SAM model, namely the recruitment $R_{y}$ ($N_{1,y}$), natural mortality $M_{a,y}$ and the fishing mortality $F_{a,y}$. We have treated these as observable processes, but in reality they can only be indirectly measured, so SAM treats them
as latent processes. The two types of measurements explicitly modeled in SAM are: 

* Catch-at-age $C_{a,y}$
* Survey index-at-age $I_{a,y}$

In this exercise we will focus on the observational equations for $C_{a,y}$ and $I_{a,y}$. 
For simplicity we treat $R_{y}$, $M_{a,y}$ 
and $F_{a,y}$ as fully observed. 

***

### Catch-at-age

The catch-at-age is predicted by the Baranov catch equation: 

\[
  C_{a,y} = \frac{F_{a,y}}{F_{a,y} + M_{a,y}} N_{a,y} (1 - e^{- F_{a,y} - M_{a,y}}).
\]
This equation is based on the assumption that the surviving part of each cohort develops over time 
accoring to the differential equation: 
\[
  \frac{dN}{dt} = -(F + M)N,
\]
which has the solution $N = N_0 e^{-F-M}$, meaning that the dead cohort size is $N_0(1 - e^{-F-M})$. 
Combined with the fact that the proportion of deaths due to fishing is $F_{a,y}/(F_{a,y} + M_{a,y})$ gives the Baranov catch equation.   

For convinience, we will define the total mortality as $Z_{a,y} = F_{a,y} + M_{a,y}$. Taking the logarithm and assuming gaussian observation noise ($\epsilon_{a,y}$), we can write our model as 
\[
  \log C_{a,y} = \log F_{a,y} - \log Z_{a,y} + \log N_{a,y} + log(1 - e^{-Z_{a,y}}) + \epsilon_{a,y},
\]
where $\epsilon_{a,y} \sim N(0,\sigma^2)$. 

### The data 

* The data list in ```Cobs.RData``` contains a vector ```Cobs``` of the catch observations. It is a
vector not a matrix.

* In addition the data list contains a matrix ```aux``` with three columns ```(year, fleet, age)```.
The $i$th row in this matrix contains the information for the $i$th element in the Cobs vector.

* Further the data list contains information on ```F, N, M, minYear```, and ```minAge```. 

*** 


### Full code for example 
<details> 
  <summary> R code </summary> 
```{r,echo=FALSE, comment=""}
include_source("Cobs.R", linesToInclud=c(1:999))
```
</details>

<details> <summary> C++ code</summary>
```{r,echo=FALSE, comment=""}
include_source("Cobs.cpp", linesToInclud=c(1:999))
```
</details>


*** 

## Including survey fleets

We will now extend the model we just implemented to include survey fleets, but also handle missing data. This will be done by adding them as random effects and estimating them. 

The survey indices are models as: 
\[
  \log I_{y}^{(s)} = \log ( Q_a^{(s)} e^{-(F_{a,y} + M_{a,y})day^{(s)}/365} N_{a,y}) + \epsilon^s_{a,y},  
\]

where 

* $Q_a^{(s)}$ is the proportionality coefficient for age $a$ and survey $s$. This will be estimated. 
* $day^{(s)}/365$ is the time into the year the survey $s$ is conducted. 

***

### The data

* The data list in `allfleets.RData` contains a vector `obs` with all observations (catches
and surveys).
* In addition the data list contains a matrix ```aux``` with three columns ```(year, fleet, age)```.
The $i$th row in this matrix contains the information for the $i$th element in the Cobs vector.
* Data also contain a vector ```fleetTypes``` with three elements. ''0'' indicates a catch fleet
and ''2'' indicates a survey fleet.
* Finally the data contains a vector `sampleTimes` with three elements, which is the $day^{(s)}/365$
values (only used for the survey fleets).

### Handling missing observations

Some of the observations in the `obs` vector are missing `NA`. We could just remove these from the likelihood, but a smarter way to deal with them is to treat them as random effects. On the R-side we add the missing values to the parameter list: 
```{r, eval = F}
missing = numeric(sum(is.na(allfleets$obs)))
```

We can make a helper function for dealing with missing values. This can be pasted right below `#include<TMB.hpp>`: 

```c++
// Function that checks if data is missing 
// Returns logical true or false
template<class Type> 
bool isNA(Type x){
  return R_IsNA(asDouble(x));
}
```

and impute the missing values: 

```c++
// count missing values 
int index_is_missing = 0; 
for(int i = 0; i < n; i++){
 // Estimate missing data
 if(isNA(obs(i))){
   obs(i) = exp(missing(index_is_missing++));
 }
}
```

## Configuring parameters 

In this data set we have 3 fleets and 9 ages, but not all fleets have all ages. We only need to focus on fleet 2 and 3, as they are the only survey fleets. If we define an integer data matrix like 

```{r, eval = F}
allfleets$keyQ <- rbind(rep(NA, 9), 
                        c(NA, 0, 1, 2, 3, 4, 5, 6, NA),
                        c(7, 8, 9, 10, 11, 12, NA, NA, NA))
```

and a parameter like 

```{r, eval = F}
logQ = numeric(max(allfleets$keyQ, na.rm = T) + 1)
```

we can us the table to look up which model parameter we should use for a given fleet $f$ and a age $a$. This can be done like: 

```c++ 
case 2:
  logPred(i) = logQ(keyQ(fleet, age)) + log(N(year, age)) - sampleTimes(fleet) * Z; 
break; 
```

### Exercise 

1. 
<details> 
  <summary> Extend the previous model for catch-at-age to include surveys. </summary> 
