---
title: "Recruitment"
author: "Hans Skaug, Jens Wahl, Olav Breivik"
date: "27 mai 2019"
output: 
  html_document
---


```{r, echo=FALSE, eval=TRUE, results='hide',include=FALSE}
library(TMB)
```

## **Recruitment**
In this example you learn: 

* The recruitment process of SAM.
* How to implement the recruitment process in TMB.
* Extract quantities from the model in R.

***
#### Recruitment as a random walk

In the SAM model we model the recruitment process ($R_y$) as an unobserved (latent) process, but we will in this example assume that we observe this process. The simplest way to model recruitment is as a random walk: 

\[
  \log R_y = \log R_{y-1} + \epsilon_y, 
\]

where $\epsilon_t \sim N(0, \sigma_R^2)$ and iid, and $R_y$ is the observed recruitment in year $y$, for $y = 1, \ldots, T$. If we assume no measurement error, the only parameter to be estimated is the standard deviation of $\sigma_R$. 

***
#### Full code for example
<details>
  <summary> R code </summary>
```{r, echo = TRUE, warning=F}
# Load data
load("Robs.RData")

# Compile TMB code for for random walk
compile("recruitment.cpp")
dyn.load(dynlib("recruitment"))

# Prepare for TMB
param <- list(log_sigma_r = -0.5)
data <- list(Robs = Robs$Robs)


# Make objective function
obj <- MakeADFun(data, param, DLL = "recruitment")
opt <- nlminb(obj$par, obj$fn, obj$gr, control = list(trace = FALSE))

# Get standard deviations
rep <- sdreport(obj)
summary(rep)

```
</details>

<details> 
  <summary> C++ code </summary>
```c++
#include <TMB.hpp>

template<class Type> 
Type objective_function<Type>::operator()(){
  
  // Data 
  DATA_VECTOR(Robs);
  
  // Parameter
  PARAMETER(log_sigma_r);

  Type sigma_r = exp(log_sigma_r);
  ADREPORT(sigma_r); // Report standard deviation of sigma_r back to R
  
  Type nll = 0; // negative log likelihood
  vector<Type> log_Robs = log(Robs); // log of recruitment
  
  // Se exercise 1 below for negative log-likelihood
  
  return(nll); 
}
```
</details> 

***

#### C++ code step-by-step 

<details> 
  <summary> Load TMB library </summary> 
```c++
#include <TMB.hpp>

template<class Type> 
Type objective_function<Type>::operator()(){

```
</details> 

<details> 
  <summary> Read data and parameters from R </summary> 
```c++
// Data 
DATA_VECTOR(Robs);

// Parameter
PARAMETER(log_sigma_r);
``` 
</details>   

<details> 
  <summary> Transform parameter </summary> 
```c+++
  Type sigma_r = exp(log_sigma_r);
  ADREPORT(sigma_r); // Report standard deviation of sigma_r back to R
```
</details> 
  
  
<details> 
  <summary> Likelihood contribution from observations </summary> 
  See exercise 1.2
</details> 

***
#### Exercice 1

1.
<details>
  <summary> Write down the negative log likelihood of the recruitment process described above.</summary>
  Since all $\epsilon_t$ are independent and $R_y$ only is dependent on $R_{y-1}$ the likelihood can be written as 
  \begin{align}
    L(\sigma_R) = \prod_{y = 2}^T N(R_{y-1}, \sigma_R^2).
  \end{align}
  The negative log likelihood is therefore given by 
  \begin{align}
    - \log L(\sigma_R) = - \sum_{y = 2}^T \log N(R_{y-1}, \sigma_R^2).
  \end{align}
</details>

2. 
<details> 
  <summary> Implement the likelihood from 1. </summary> 
  
    ```c++ 
      for(int i = 1; i < Robs.size(); i++){
        nll -= dnorm(log_Robs(i), log_Robs(i - 1), sigma_r, true);
      }

    ```
  
</details> 

### Recruitment as a state space model 

Assume that we don't observe the recruitment process directly and that measurement error is introduced. We assume that the measurment error $\eta_y \sim N(0, \sigma_{R^\text{obs}}^2)$ and iid. The model can then be expressed as:

\begin{align}
  \log R_y^{(\text{obs})} &= R_y + \eta_y, \\
  \log R_y &= \log R_{y-1} + \epsilon_y. 
\end{align}

The "true" process, $R_y$, is now unobserved and need to be integrated out. 

#### Full code for example
<details> 
  <summary> R code </summary> 
```{r non.finished.plotting, echo=TRUE, comment = "", message=FALSE, results = "hide", warning=FALSE}
# Compile TMB code for random walk state space model
compile("recruitment_state_space.cpp")
dyn.load(dynlib("recruitment_state_space"))

param <- list(logR = rep(0, length(Robs$Robs)),
              log_sigma_Robs = -0.5,
              log_sigma_logR = -0.5)

data <- list(Robs = Robs$Robs)

obj <- MakeADFun(data, param, random = "logR", DLL = "recruitment_state_space")
opt <- nlminb(obj$par, obj$fn, obj$gr)

# Get standard deviation of parameters and latent process
rep <- sdreport(obj)

# Get summary of fixed effects 
summary(rep, "fixed", p.value = T)
logR_rep <- summary(rep, "random")


# Plot estimated recruitment process
{plot(Robs$year, log(Robs$Robs), ylab = "Recruitment", xlab = "year")
lines(Robs$year, logR_rep[, 1], col="black")
lines(Robs$year, logR_rep[, 1] - 2 * logR_rep[, 2], col="black", lty="dashed")
lines(Robs$year, logR_rep[, 1] + 2 * logR_rep[, 2], col="black", lty="dashed")}
```

</details>

<details> 
  <summary> C++ code </summary>
```c++
#include <TMB.hpp>

template<class Type> 
Type objective_function<Type>::operator()(){
  
  DATA_VECTOR(Robs); 
  
  PARAMETER_VECTOR(logR); // Latent process 
  PARAMETER(log_sigma_Robs); 
  PARAMETER(log_sigma_logR); 
  
 // Transform data
 vector<Type> log_Robs = log(Robs);
  
 // Transform standard deviations
 Type sigma_Robs = exp(log_sigma_Robs); 
 Type sigma_logR = exp(log_sigma_logR);
 
 ADREPORT(sigma_Robs); 
 ADREPORT(sigma_logR); 
 
 // Negative log likelihood
 Type nll = 0; 
 // Number of observations 
 Type n = Robs.size(); 
 
 // Contribution to likelihood from latent process logR
 for(int i = 1; i < n; i++){
   nll -= dnorm(logR(i), logR(i - 1), sigma_logR, true);
 }
 
 // Contribution to likelihood from observations Robs
  
 // See exercise 2.2 below
 
 return(nll);
  
}
```
    
</details> 



#### C++ code step-by-step 

<details> 
  <summary> Load TMB library </summary> 
```c++
  #include <TMB.hpp>
  
  template<class Type> 
  Type objective_function<Type>::operator()(){

```
</details> 

<details> 
  <summary> Read data and parameters from R </summary> 
```c++
  DATA_VECTOR(Robs); 
  
  PARAMETER_VECTOR(logR); // Latent process 
  PARAMETER(log_sigma_Robs); 
  PARAMETER(log_sigma_logR); 
  
  // Transform data
  vector<Type> log_Robs = log(Robs);
``` 
</details>   

<details> 
  <summary> Transform parameters </summary> 
```c+++
 // Transform standard deviations
 Type sigma_Robs = exp(log_sigma_Robs); 
 Type sigma_logR = exp(log_sigma_logR);
 
 ADREPORT(sigma_Robs); 
 ADREPORT(sigma_logR); 
```
</details> 
  
  
<details> 
  <summary> Likelihood contribution from latent process and observations</summary> 
```c++
 // Contribution to likelihood from latent process logR
 for(int i = 1; i < n; i++){
   nll -= dnorm(logR(i), logR(i - 1), sigma_logR, true);
 }
 
 // Contribution to likelihood from observations Robs
  
 // See exercise 2.2 below
 
 return(nll);
  
}
```
</details> 


***
#### Exercise 2

1. 
<details> 
  <summary> Write down the joint negative log likelihood of the recruitment state-space model </summary> 
  Let $R^{\text(obs)} = (R_1^\text{obs}, \ldots, R_T^\text{obs})$ and $R = (R_1, \ldots, R_T)$. The joint distribution of        $(R^\text{obs}, R)$ can be written as 
  \begin{align}
    L(\theta) &= \int f(R^\text{obs}, R) dR = \int f_{R^\text{obs}}(R^\text{obs} | R) f_R(R) dR \\
    &= \int f_{R_1^\text{obs}}(R_1^\text{obs} | R_1) f_{R_1}(R_1) \prod_{y=2}^T f_{R_y^\text{obs}}(R_y^\text{obs} | R_y) f_{R_y}(R_y) dR,
  \end{align}
  where $f_{R_y^\text{obs}}$ is the conditional density of $\log R_y^\text{obs}$ given $R_y$ and $f_{R_y}$ is the density of the latent recruitment $R_y$.
  The negative log likelihood we need to implement in TMB is therefore 
  \[
    - \log L(\theta) = -N(R_1,\sigma_{R^\text{obs}}^2) - \sum_{y=2}^T N(R_y,\sigma_{R^\text{obs}}^2) - \sum_{y=2}^T \log             N(R_{y-1}, \sigma_R^2),
  \]
  where the two first terms are the contribution from the observations, while the last sum is the likelihood contribution from   the unobserved process. 
  
</details> 

2. 
<details> 
  <summary> Implement the likelihood for the observation equation. </summary>
    ```c++
       // Contribution to likelihood from observations Robs
       for(int i = 0; i < n; i++){
         nll -= dnorm(log_Robs(i), logR(i), sigma_Robs, true);
       }
    ```
</details> 

3. 
<details> 
  <summary> Plot the estimated recruitment process with 2 times standard error </summary> 
    ```r
    plot(Robs$year, log(Robs$Robs), ylab = "Recruitment", xlab = "year")
    lines(Robs$year, logR_rep[, 1], col="black")
    lines(Robs$year, logR_rep[, 1] - 2 * logR_rep[, 2], col="black", lty="dashed")
    lines(Robs$year, logR_rep[, 1] + 2 * logR_rep[, 2], col="black", lty="dashed")
    ```
</details> 



***
### Using spawning stock biomass (SSB) to predict recruitment.

Instead of letting the recruitment be a function of the recruitment from the year before, we can use the spawning stock biomass from last year to predict the recruitment. 
We will look at two recruitment functions: 

* Richer: $f(SSB) = \alpha\cdot SSB \cdot e^{-\beta\cdot SSB}$
* Beverton-Holt: $f(SSB) = \frac{\alpha\cdot SSB}{1 + \beta\cdot SSB}$

We can then setup our recruitment process like: 

\[
  \log R_y = \log f(SSB_{y - 1}) + \epsilon_y, \quad \text{where } \epsilon \sim N(0, \sigma_R^2)
\]



*** 
#### Full code for example (Ricker)

<details> 
  <summary> R code </summary> 
```{r, echo=-1, comment = "", message=FALSE, results = "hide", warning=FALSE}

# Richer 

compile("recruitment_state_space_SSB.cpp")
dyn.load(dynlib("recruitment_state_space_SSB"))

data <- list(Robs = Robs$Robs,
             SSB = Robs$ssb,
             method = 1)

param <- list(logR = rep(0, length(Robs$Robs)),
              log_sigma_Robs = 0,
              log_sigma_logR = 0, 
              ricker = if(data$method == 1){numeric(2)} else{numeric(0)},
              bh = if(data$method == 2){numeric(2)} else{numeric(0)})



obj <- MakeADFun(data, param, random = "logR", DLL = "recruitment_state_space_SSB")
opt <- nlminb(obj$par, obj$fn, obj$gr)
rep <- sdreport(obj)
logR_r_rep <- summary(rep, "random")


# Beverton-Holt


data <- list(Robs = Robs$Robs,
             SSB = Robs$ssb,
             method = 2)

param <- list(logR = rep(0, length(Robs$Robs)),
              log_sigma_Robs = 0,
              log_sigma_logR = 0, 
              ricker = if(data$method == 1){numeric(2)} else{numeric(0)},
              bh = if(data$method == 2){numeric(2)} else{numeric(0)})



obj <- MakeADFun(data, param, random = "logR", DLL = "recruitment_state_space_SSB")
opt <- nlminb(obj$par, obj$fn, obj$gr)
rep <- sdreport(obj)
logR_bh_rep <- summary(rep, "random")





# Plot all estimated recruitment processes
{plot(Robs$year, log(Robs$Robs), ylab = "Recruitment", xlab = "year")
lines(Robs$year, logR_rep[, 1], col="black")
lines(Robs$year, logR_rep[, 1] - 2 * logR_rep[, 2], col="black", lty="dashed")
lines(Robs$year, logR_rep[, 1] + 2 * logR_rep[, 2], col="black", lty="dashed")
lines(Robs$year, logR_r_rep[, 1], col = "red")
lines(Robs$year, logR_bh_rep[, 1], col = "blue")
legend("bottomright", c("Random walk", "Ricker", "Berverton-Hold"), lty = 1, col = c("black", "red", "blue"))}


```

</details>

<details> 
  <summary> C++ code </summary>
```c++
#include <TMB.hpp>

template<class Type> 
Type objective_function<Type>::operator()(){
  
  DATA_VECTOR(Robs);
  DATA_VECTOR(SSB);
  DATA_INTEGER(method);
  
  PARAMETER_VECTOR(logR); // Latent process 
  PARAMETER(log_sigma_Robs); 
  PARAMETER(log_sigma_logR); 
  PARAMETER_VECTOR(ricker); 
  PARAMETER_VECTOR(bh); 
  
  // Transform data
  vector<Type> log_Robs = log(Robs);
  
 // Transform standard deviations
 Type sigma_Robs = exp(log_sigma_Robs); 
 Type sigma_logR = exp(log_sigma_logR);
 
 ADREPORT(sigma_Robs); 
 ADREPORT(sigma_logR); 
 
 // Negative log likelihood
 Type nll = 0; 
 // Number of observations 
 Type n = Robs.size(); 
 Type pred = 0; 
 // Contribution to likelihood from latent process logR
 for(int i = 1; i < n; i++){
   
   switch(method){
   case 0:
     pred = logR(i - 1);
     break;
     
   case 1:
     pred = ricker(0) + log(SSB(i - 1)) - exp(ricker(1)) * SSB(i - 1);
     break;

   case 2:
     // See exercice below
     break;

   default:
     std::cout << "This method is not implementet. Method has to be a number 0, 1, 2" << std::endl;
   break;
   }
   
   nll -= dnorm(logR(i), pred, sigma_logR, true);

 }
 
 // Contribution to likelihood from observations Robs
 for(int i = 0; i < n; i++){
   nll -= dnorm(log_Robs(i), logR(i), sigma_Robs, true);
 }
 
 return(nll);
  
}
```
    
</details> 

***
### Exercise 3

1. 
<details> 
  <summary> Implement the Beverton-Holt model in the TMB program (note that the parameters are on log scale) 
  </summary> 

    ```c++
        case 2:
         pred = bh(0) + log(SSB(i - 1)) - log(Type(1.0) + exp(bh(1)) * SSB(i - 1));
         break;
    ``` 
</details> 


  
  