---
title: "Survival"
author: "Hans Skaug, Jens Wahl, Olav Breivik"
date: "6 juni 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(TMB)
source("../../utils.R")

```

## **Survival**

In the first example we looked at the recruitment process, where we modeled the number of fish
born ($R_y$) in year $y$. The next step is to look at survival of individulas aged $a$ 
from one year to the next, i.e. age-structured population dynamics. Denote by $N_{a,y}$ the number
of fish aged $a$ alive in year $y$. For example, $N_{1, 8}$ is the number of fish of age 1 in year 8 (i.e $R_8$). The number of fish next year equals the number of fish this year minus those who have died and added those who have been born. This can be expressed mathematically as:

\begin{align}
\log N_{a,y} = \log N_{a-1,y-1} - F_{a-1,y-1} - M_{a-1, y-1} +  \epsilon_{ay}, \quad \text{where }  \epsilon_{ay} \sim  N(0, \sigma_S^2)
\end{align}

where $F$ is the fishing mortality and $M$ is the natural mortality. If our model contains a maximum age group $A$, then the survival for age $A$ is given by: 

\begin{equation}
  \label{eq:survival}
  \log N_{A,y} = \log (N_{A-1,y-1} e^{- F_{A-1,y-1} - M_{A-1, y-1}} + N_{A,y-1} e^{ - F_{A,y-1} - M_{A,   y-1}}) + \eta_{Ay}, 
\end{equation}
where the contribution is the survival form last year in the age group below, but also from the same age group from last year, since this is the maximum age. 

*** 

### Survival as state space model 

As we did with the recruitment process, we will model the survival as a state space model in the following way: 

\begin{align}
  \log N_{ay}^{(\text{obs})} &= \log N_{a,y} + \epsilon_{ay}, \quad \text{where }\epsilon_{ay} \sim N(0, \sigma^2) \\
  \log N_{a,y} &= \log N_{a-1,y-1} - F_{a-1,y-1} - M_{a-1,y-1} + \eta_{ay}\\
 \log N_{A,y} &= \log (N_{A-1,y-1} e^{- F_{A-1,y-1} - M_{A-1, y-1}} + N_{A,y-1} e^{ - F_{A,y-1} - M_{A, y-1}}) + \eta_{Ay}
 \end{align}

Note that for $a = 1$ we have the recruitment process from last exercise. 

***

### Full code for example 

<details> 
  <summary> R code </summary> 
```{r, echo = TRUE, warning=F, results='hide'}

load("Nobs.RData")
matplot(log(Nobs$Nobs), main="logN")

compile("survival.cpp")
dyn.load(dynlib("survival"))

# Prepare for TMB
param <- list(logN = matrix(0 ,nrow = nrow(Nobs$Nobs), ncol = ncol(Nobs$Nobs)),
              log_sigma_Nobs = 0, 
              log_sigma_logN = 0,
              log_sigma_logR = 0)

data <- list(Nobs = Nobs$Nobs,
             F = Nobs$F,
             M = Nobs$M)

obj <- MakeADFun(data, param, random = "logN", DLL = "survival")
opt <- nlminb(obj$par, obj$fn, obj$gr, control = list(trace = 0))

rep <- sdreport(obj)

logN_rep <- summary(rep, "random")

# Add lines for estimated survival for all age groups
matplot(as.list(rep ,"Est")$logN, type="l", add=TRUE)

```
  
</details> 

<details> 
  <summary> C++ code </summary> 
```c++
#include <TMB.hpp>

template<class Type> 
Type objective_function<Type>::operator()(){
  
  DATA_MATRIX(Nobs); 
  DATA_MATRIX(M); 
  DATA_MATRIX(F); 
  
  PARAMETER_MATRIX(logN); // Survival matrix for all age groups and years 
  PARAMETER(log_sigma_Nobs); // standard deviation for observations 
  PARAMETER(log_sigma_logN); // standard devitation for state equation
  PARAMETER(log_sigma_logR); // stadard deviation for recruitment 
  
  // Transform parameters
  Type sigma_Nobs = exp(log_sigma_Nobs); 
  Type sigma_logN = exp(log_sigma_logN); 
  Type sigma_logR = exp(log_sigma_logR);
  
  // Report to R
  ADREPORT(sigma_logN);
  ADREPORT(sigma_Nobs);
  ADREPORT(sigma_logR);
  

  
  Type nll = 0; 
  
  // Number of years and number of age groups
  int n_year = Nobs.rows();
  int n_age = Nobs.cols(); 
  
  // Recruitment for age zero
  for(int y = 1; y < n_year; y++){
    nll -= dnorm(logN(y, 0), logN(y - 1, 0), sigma_logR, true);
  }
  
  // survival for age greater than zero
  
  // Contribution form latent process
  for(int y = 1; y < n_year; y++){
    for(int a = 1; a < n_age; a++){
      
      // If we are at max age we get contribution from same group last year
      if(a == n_age - Type(1)){
        Type pred = log(exp(logN(y - 1, a - 1) - F(y - 1, a - 1) - M(y - 1, a - 1))) +
                     log(exp(logN(y - 1, a) - F(y - 1, a) - M(y - 1, a)));
        nll -= dnorm(logN(y, a), pred, sigma_logN, true);
      } else{
       // see exercise below for implementation of N_ay
      }
    }
  }
  
  // Contribution from observations given latent process
  
  for(int y = 0; y < n_year; y++){
    for(int a = 0; a < n_age; a++){
     nll -= dnorm(log(Nobs(y, a)), logN(y, a), sigma_Nobs, true);
    }
  }
  
  return nll;
  
}
``` 

*** 

### C++ code step-by-step 
<details> 
  <summary> Read data and parameters from R </summary>
```c++
PARAMETER_MATRIX(logN); // Survival matrix for all age groups and years 
PARAMETER(log_sigma_Nobs); // standard deviation for observations 
PARAMETER(log_sigma_logN); // standard devitation for state equation
PARAMETER(log_sigma_logR); // stadard deviation for recruitment 

// Transform parameters
Type sigma_Nobs = exp(log_sigma_Nobs); 
Type sigma_logN = exp(log_sigma_logN); 
Type sigma_logR = exp(log_sigma_logR);

// Report to R
ADREPORT(sigma_logN);
ADREPORT(sigma_Nobs);
ADREPORT(sigma_logR);
```
</details> 

<details> 
  <summary> Contribution from $N_{1,y}$ </summary>
```c++
// Number of years and number of age groups
Type nll = 0; 
int n_year = Nobs.rows();
int n_age = Nobs.cols(); 

// Recruitment for age zero
for(int y = 1; y < n_year; y++){
  nll -= dnorm(logN(y, 0), logN(y - 1, 0), sigma_logR, true);
}
``` 
</details> 

<details> 
  <summary> Contribution from $N_{a,y}$ for $a > 1$ </summary> 
```c++
// survival for age greater than zero

// Contribution form latent process
for(int y = 1; y < n_year; y++){
  for(int a = 1; a < n_age; a++){
    
    // If we are at max age we get contribution from same group last year
    if(a == n_age - Type(1)){
      Type pred = log(exp(logN(y - 1, a - 1) - F(y - 1, a - 1) - M(y - 1, a - 1))) +
                   log(exp(logN(y - 1, a) - F(y - 1, a) - M(y - 1, a)));
      nll -= dnorm(logN(y, a), pred, sigma_logN, true);
    } else{
       // see exercise below for implementation of N_ay
    }
  }
}
```
</details> 

<details> 
  <summary> Contribution form observations $N_{a,y}^{\text{obs}}$ </summary> 
```c++
// Contribution from observations given latent process

for(int y = 0; y < n_year; y++){
  for(int a = 0; a < n_age; a++){
   nll -= dnorm(log(Nobs(y, a)), logN(y, a), sigma_Nobs, true);
  }
}

return nll;

}

```

*** 

### Exercise 1

1. 
<details> 
  <summary> Implement the contribution from $N_{a,y}$ for $1 < a < A$. This is the second equation above. Use the code for $N_{A,y}$   </summary> 
    ```c++
        nll -= dnorm(logN(y ,a), logN(y - 1, a - 1) - F(y - 1, a - 1) - M(y - 1, a - 1), sigma_logN, true);
    ```
</details> 


*** 

## Fishing mortality 

The last latent process in SAM is the fishing mortality $F_{a,y}$ for age $a$ in year $y$. As with the survival process, we will model this as a random walk with observational error, but we will also have add a correlation structure in the year dimension. If we let ${\bf F}_y = (F_{1,y}, \ldots, F_{A,y})$, we can formulate a model for $F_y$ as

\[
  \log {\bf F}_y = \log {\bf F}_{y-1} + \pmb{\xi_y},
\]

where $\pmb{\xi_y}$ follows an A (maximum age) -dimensional multivariate normal distribution with expectation 0 and covariance matrix $\Sigma$. There are many options for the structure of $\Sigma$, but we will focus on three alternatives: 

1. Age groups are independent: $\Sigma_{ij} = 0$ for year $i \ne j$.
2. Same correlation between all age groups (compound symmetry): $\Sigma_{ij} = \rho \sqrt{\Sigma_{ii} \Sigma_{jj}}$.
3. AR(1) structure with exponentially decaying correlations: $\Sigma_{ij} = \rho^{|i-j|}\sqrt{\Sigma_{ii} \Sigma_{jj}}.$

As with the survival $N_{a,y}$, we observe 
\[
  \log F_{a,y}^\text{obs} = \log F_{a,y} + \eta_{ay}, \quad \eta_{ay} \sim N(0, \sigma_{F^\text{obs}}^2),
\]

and have the following state-space model: 

\begin{align}
    \log F_{a,y}^\text{obs} &= \log F_{a,y} + \eta_{ay}, \\ 
  \log {\bf F}_y &= \log {\bf F}_{y-1} + \pmb{\xi_y} 
\end{align}


*** 

### Full code for example 
<details> 
  <summary> R code </summary> 
```{r,echo=FALSE, comment=""}
include_source("Fobs.R", linesToInclud=c(1:999))
```
</details>

<details> 
  <summary> C++ code </summary> 
```c++
#include<TMB.hpp>

// Helper function to ensure valid correlation 
template<class Type> 
Type trans(Type trans_rho){
  Type rho = (exp(trans_rho) - Type(1)) / (exp(trans_rho) + Type(1));
  return rho;
}

template<class Type> 
Type objective_function<Type>::operator()(){
  
  DATA_MATRIX(Fobs);
  DATA_INTEGER(cormode); // Correlation structure
  
  PARAMETER_MATRIX(logF); // Fishing mortalitu for all ages and years 
  PARAMETER_VECTOR(log_sigma_logF); // SD for all ages
  PARAMETER_VECTOR(trans_rho); // Correlation
  PARAMETER(log_sigma_Fobs); // SD for observations 
  
  
  // Number of years and age groups
  int n_year = Fobs.rows();
  int n_age = Fobs.cols();
  
  // Transform parameters 
  Type sigma_Fobs = exp(log_sigma_Fobs); 
  vector<Type> sigma_logF = exp(log_sigma_logF); 
  Type rho = 0;
  

  // Report standard deviation of parameters
  ADREPORT(sigma_Fobs);
  ADREPORT(sigma_logF);

  
  // Covariance matrix for age groups 
  matrix<Type> Sigma(n_age, n_age);
  Sigma.setZero();
  
  // Negative log likelihood
  Type nll = 0; 
  
  
  switch(cormode){
  
    // Independent
    case 0:
      Sigma.diagonal() = sigma_logF * sigma_logF;
    break;
      
    // Compound symmertry
    case 1:
      Sigma.diagonal() = sigma_logF * sigma_logF;
      rho = trans(trans_rho(0)); 
      for(int i = 0; i < n_age; i++){
        for(int j = 0; j < i; j++){
          Sigma(i, j) = rho * sigma_logF(i) * sigma_logF(j);
          Sigma(j, i) = Sigma(i, j);
        }
      }
    break;
      
    // AR(1) 
    case 2: 
      // See exercise below for implementation of AR(1) structure
    break;
      
    default:
      std::cout<<"Error: This cormode not implemented yet."<<std::endl;
      exit(EXIT_FAILURE);
    break;
  }
  
  // Contribution from latent process
  
  // Make zero mean multivariate object with covariance Sigma
  // negative_mvn will return the negative log of the multivariate normal density
  density::MVNORM_t<Type> negative_mvn(Sigma);
  ADREPORT(Sigma);
  
  
  for(int y = 1; y < n_year; y++){
    nll += negative_mvn(logF.row(y) - logF.row(y - 1)); // Returns negative log likelihood  
  }
  
  // Contribution from observations
  for(int y = 0; y < n_year; y++){
    for(int a = 0; a < n_age; a++){
      nll -= dnorm(log(Fobs(y, a)), logF(y, a), sigma_Fobs, true);
    }
  }
  
  return nll;
  
}
```

*** 

### Exercise 2
1.
<details> 
  <summary> Implement the AR(1) correlation structure. Tips: Use ```Sigma.diagonal()```to fill the diagonal     of the covariance matrix. </summary> 
    ```c++ 
        Sigma.diagonal() = sigma_logF * sigma_logF;
        rho = trans(trans_rho(0)); 
        for(int i = 0; i < n_age; i++){
          for(int j = 0; j < i; j++){
            Sigma(i, j) = pow(rho, Type(i - j)) * sigma_logF(i) * sigma_logF(j);
            Sigma(j, i) = Sigma(i, j);
          }
        }
    ```
</details> 


2. 
<details> 
  <summary> Compare the models with different correlation structure and find the best one based on AIC.  </summary> 
    ```{r, eval=T, results='hide'}
    compile("Fobs.cpp")
    dyn.load(dynlib("Fobs"))
    
    
    load("Fobs.Rdata")
    
    # Calculate AIC for a given model
    TMBAIC <- function(opt){
        k <- length(opt$par)
        aic <- 2 * k + 2 * opt$objective
        return(aic)
      }
    
    
    # No correlation - Model 1 -----
    data <- list(Fobs = Fobs$Fobs,
                 cormode = 0)
    
    param <- list(logF = matrix(0, nrow = nrow(Fobs$Fobs), ncol = ncol(Fobs$Fobs)),
                  log_sigma_logF = rep(0, ncol(Fobs$Fobs)),
                  trans_rho = if(data$cormode == 0){numeric(0)} else{0.1},
                  log_sigma_Fobs = 0)
    
    obj_mod0 <- MakeADFun(data, param, random = "logF", DLL = "Fobs")
    opt_mod0 <- nlminb(obj_mod0$par, obj_mod0$fn, obj_mod0$gr)
    
    # Plot the estimated fish mortality for all years 
    rep <- sdreport(obj_mod0)
    matplot(Fobs$year, log(Fobs$Fobs), xlab="Year", ylab="logF", pch=colnames(Fobs$Fobs), main = "Model 1")
    matplot(Fobs$year, as.list(rep, "Est")$logF, type="l", add=TRUE)
    
    
    # Compund symmetry - Model 2 ------
    data <- list(Fobs = Fobs$Fobs,
                 cormode = 1)
    
    param <- list(logF = matrix(0, nrow = nrow(Fobs$Fobs), ncol = ncol(Fobs$Fobs)),
                  log_sigma_logF = rep(0, ncol(Fobs$Fobs)),
                  trans_rho = if(data$cormode == 0){numeric(0)} else{0.1},
                  log_sigma_Fobs = 0)
    
    obj_mod1 <- MakeADFun(data, param, random = "logF", DLL = "Fobs")
    opt_mod1 <- nlminb(obj_mod1$par, obj_mod1$fn, obj_mod1$gr)
    
    # Plot the estimated fish mortality for all years 
    rep <- sdreport(obj_mod1)
    matplot(Fobs$year, log(Fobs$Fobs), xlab="Year", ylab="logF", pch=colnames(Fobs$Fobs), main = "Model 2")
    matplot(Fobs$year, as.list(rep, "Est")$logF, type="l", add=TRUE)
    
    # AR(1) - Model 3 ------
    data <- list(Fobs = Fobs$Fobs,
                 cormode = 2)
    
    param <- list(logF = matrix(0, nrow = nrow(Fobs$Fobs), ncol = ncol(Fobs$Fobs)),
                  log_sigma_logF = rep(0, ncol(Fobs$Fobs)),
                  trans_rho = if(data$cormode == 0){numeric(0)} else{0.1},
                  log_sigma_Fobs = 0)
    
    obj_mod2 <- MakeADFun(data, param, random = "logF", DLL = "Fobs")
    opt_mod2 <- nlminb(obj_mod2$par, obj_mod2$fn, obj_mod2$gr)
    
    # Plot the estimated fish mortality for all years 
    rep <- sdreport(obj_mod2)
    matplot(Fobs$year, log(Fobs$Fobs), xlab="Year", ylab="logF", pch=colnames(Fobs$Fobs), main = "Model 3")
    matplot(Fobs$year, as.list(rep, "Est")$logF, type="l", add=TRUE)
    
    # Calculate AIC for all models - we want the smallest
    TMBAIC(opt_mod0) # 895
    TMBAIC(opt_mod1) # 565
    TMBAIC(opt_mod2) # 478 
    
    ```
</details> 